Homoiconicity Without Template Haskell

### 2016-06-10 paper/paper.tex
more paper edits

\usepackage{amsmath,amsfonts}
\usepackage{url}

@@

\begin{filecontents}{paper.bib}
@article{goldberg1991,
  title={What every computer scientist should know about floating-point arithmetic},
  author={Goldberg, David},
  journal={ACM Computing Surveys (CSUR)},
  volume={23},
  number={1},
  pages={5--48},
  year={1991},
  publisher={ACM}
}

@@

@misc{viewpatterns,
  author = {GHC Wiki},
  title = {View patterns: lightweight views for Haskell},
  url = {https://ghc.haskell.org/trac/ghc/wiki/ViewPatterns},
  note = {Accessed: 2016-06-09}
}

@@

%\title{What's the point of homoiconicity?}
%\title{The (floating) point of homoiconicity}
\title{Homoiconicity Without Template Haskell}
\subtitle{Functional Pearl}

@@

\begin{abstract}
We say a function is \emph{homoiconic} if it can be converted into an isomorphic abstract syntax tree (AST).
This AST can then be manipulated arbitrarily and converted back into a function.
Template Haskell gives a \emph{compile time} form of homoiconicity to Haskell programs;
but in this pearl, we demonstrate another form of homoiconicity that relies on polymorphism and GHC's type programming capabilities.
To motivate our technique,
we construct a small library that automatically transforms numerically unstable code into stable code.

%This pearl shows that many Haskell functions are \emph{homoiconic}.
%That is, we can convert these functions into isomorphic abstract syntax trees (ASTs),
%and these ASTs can be converted back into their original Haskell functions.
%%These ASTs are ordinary Haskell data types,
%%so they can be manipulated arbitrarily to perform program transformations.
%The construction does not require template Haskell,
%and so works on expressions available at either compile or run time.
%To motivate this homoiconicity,
%we construct a small library that automatically transforms numerically unstable code into stable code.

%Floating point numbers are dangerous.
%Due to their limited precision,
%seemingly correct mathematical formulae can give wildly inaccurate results.
%This pearl shows that Haskell's type system can mitigate this danger.
%
%The main observation is that many numerical Haskell functions are \emph{homoiconic}.
%That is, we can easily convert them into isomorphic abstract syntax trees (ASTs).
%The construction does not use template Haskell, and so works at both compile and run time.
%We will use these ASTs to solve two problems with floating point numbers:
%\begin{enumerate}
%\item
%We will write higher order functions that automatically stabilize (or optimize!) mathematical formulae.
%%This lets one programmer write naive floating point code;
%%then an expert on numerical analysis can write code that automatically improves the naive code.
%\item
%We will specify type class laws for floating point arithmetic that can be automatically tested.
%Existing test frameworks do not work on floating point numbers because floats do not obey many traditional algebraic laws like associativity.
%\end{enumerate}
%We demonstrate these techniques using GHC's built-in numeric class hierarchy,
%but these techniques generalize to user defined class hierarchies as well.
\end{abstract}

@@

In this pearl, we see how to write functions like |stabilize| that perform program transformations.
These functions follow the following pattern.
First, convert the input function into an abstract syntax tree (AST).
Next, manipulate the AST to remove the numerical instabilities.
Finally, convert the AST back into a callable function.
%The key observation is that certain Haskell functions are \emph{homoiconic}.
%That is, we can convert them into isomorphic abstract syntax trees (ASTs).
%The key observation is that certain Haskell functions (which we call \emph{homogeneous} and \emph{heterogeneous}) can be converted into equivalent abstract syntax trees (ASTs).
%The |stabilize| function converts its input function into an AST,
%then manipulate that AST to remove numerical stabilities,
%and finally convert the AST back into a callable function.

@@

@@

In this section we first define homogeneous type classes and functions.
There are two intuitions for homogeneous type classes.
First, they have ASTs that contain only a single type variable.
Second, we can construct an FAlgebra for these type classes.

@@

\begin{defn}
We call a function homogeneous if:
\begin{enumerate}
\item
there is exactly one type variable in the signature, which we denote by |a|;
\item
each of the function's parameters is either
\begin{enumerate}
\item
|a|, or
\item
concrete;
\end{enumerate}
\item
the function's return type is |a|; and
\item
the constraints contain only homogeneous type classes applied to |a|.
\end{enumerate}
The |logLogistic1| function is homogeneous,
and Figure \ref{code:ghc} shows many more examples taken from the Prelude's numeric hierarchy.
\end{defn}

@@

The data family |Sig| encodes what mathematicians call the \emph{signature} of the FAlgebra.
The signature defines all the operations that can be performed on an FAlgebra.
Each |Sig| data instance can have many constructors,
and these constructors come in two flavors.
First is the \emph{function constructor}.
There should be one function constructor for each homogeneous class function, and
this constructor should have the same parameters as the function.
For the |Fractional| example, we've defined the three constructors |Sig_div|, |Sig_recip|, and |Sig_fromRational| for the three member functions |(/)|, |recip|, and |fromRational| respectively.
The second constructor flavor is the \emph{superclass constructor}.
There should be one superclass constructor for each superclass, and
this constructor should store the |Sig| of the corresponding superclass.
For the |Fractional| example, we have the |Sig_Fractional_Num| constructor corresponding to the |Num| superclass.

@@

@@

Using our uniform representation of an AST,
we can make the |AST| type an instance of the homogeneous type classes.
As an example, the |Fractional| instance is shown below.
\begin{lstlisting}
instance View Fractional alg
    => Fractional (AST alg a) where
    (/) e1 e2 = Free $ embedSig $ Sig_div e1 e2
    recip e   = Free $ embedSig $ Sig_recip e
    fromRational r
        = Free $ embedSig $ Sig_fromRational r
\end{lstlisting}
For each function in the type class,
we simply embed the function's constructor into the AST by prepending the expression |Free $ embedSig $|.

We can easily create ASTs using these instances.
The |Num| AST can be created with the expression
\begin{lstlisting}
expr4 :: AST Num Double
expr4 = (1+2)+3
\end{lstlisting}
And the |Floating| AST can be created by just changing the type signature.
\begin{lstlisting}
expr5 = AST Floating Double
expr5 = (1+2)+3
\end{lstlisting}
Notice that in these last two examples we did not embed the numbers into the AST with the |Pure| constructor.
Since |AST| is an instance of |Num|,
GHC will call |fromInteger| to embed the number for us automatically.

@@

\subsection{Showing functions}
Reconsider our |logLogistic1| function.
It has type
\begin{lstlisting}
logLogistic1 :: Floating a => a -> a
\end{lstlisting}
What happens if we specialize the |a| parameter to |AST Floating Double|?
We get a peek at the function's internals.
\begin{lstlisting}
ghci> logLogistic1 1 :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (fromInteger 1))))))
\end{lstlisting}
We don't just have to input single numbers;
we can also input full expressions.
These expressions will also get displayed,
intermixed with the original function.
\begin{lstlisting}
ghci> logLogistic1 (1*3+4) :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (((fromInteger 1)*(fromInteger 3))
    +(fromInteger 4)))))))
\end{lstlisting}
And we can even display the resulting AST of nested function calls.
\begin{lstlisting}
ghci> logLogistic1 (logLogistic1 1)
    :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (log ((fromInteger 1)/((fromInteger 1)
    +(exp (negate (fromInteger 1)))))))))))
\end{lstlisting}
It would be nice if the displayed function would display the actual input variables,
rather than just substituting numbers.
To accomplish this, we need a variable type.
\begin{lstlisting}
newtype Var = Var String
\end{lstlisting}
Values of type |Var| cannot be plugged directly into the |logLogistic1| function because |Var| is not an instance of |Floating|.
Fortunately, we have a way to make any type an instance of |Floating| via the |AST Floating| type.
More generically, we define the following variables.
\begin{lstlisting}
var1 :: AST alg Var
var1 = Pure "var1"

@@

var3 :: AST alg Var
var3 = Pure "var3"
\end{lstlisting}
And we can use them to recover the exact AST of a function.
\begin{lstlisting}
ghci> logLogistic1 var1 :: AST Floating Var
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate var1)))))
\end{lstlisting}
At this point, you might be tempted to write a |Show| instance for functions.
\begin{lstlisting}
instance
    ( FAlgebra alg
    , Show (Sig alg (Free (Sig alg) Var))
    ) => Show (AST alg Var -> AST alg Var) where
    show f = show (f var1)
\end{lstlisting}
Unfortunately, if you try to use the instance naively
\begin{lstlisting}
ghci> logLogistic1
\end{lstlisting}
then ghci will complain about ambiguous type variables.
GHC's defaulting mechanism is not powerful enough for it to properly specialize the function without a type signature.
So unfortunately this technique doesn't give us a convenient |Show| instance,
it does give us a slightly more verbose alternative for showing a function.

@@

@@

We're now ready to write program transformations that manipulate the ASTs.
The standard way to deconstruct an AST Haskell is via pattern matching.
In this section, we'll first demonstrate how to pattern match our ASTs using the |ViewPatterns| and |PatternSynonyms| language extensions.
Then we'll see three simple examples of program transformations implemented in this way.

@@

Consider the following simple,
but numerically unstable function.
\begin{lstlisting}
testFunc1 :: Floating a => a -> a
testFunc1 = log (exp a)
\end{lstlisting}
As in our |logLogistic1| function, |testFunc| gives incorrect answers when specialized to |Double|.
Inputs greater than (approximately) 710 cause the |exp| function to overflow to infinity;
the log of infinity is still infinity;
so the returned answer is infinity.
But this wrong.
Mathematically, the $\log$ function is defined to be the inverse of $\exp$,
so the final result should be exactly the same as the input.
We'll write a program transformation that stabilizes the |testFunc1| function,
then we'll demonstrate how the |ViewPatterns| and |PatternSynonyms| language extensions simplify this transformation.
\label{sec:viewpatterns}

Our goal is to take in input AST,
find all occurrences of the expression |log (exp x)|,
and replace them with just |x|.
Here |x| can be an arbitrary subtree of the original AST.
We will create a series of four functions to accomplish this task,
each improving on the previous.
Here is the first.
\begin{lstlisting}
logexpAST1 :: AST Floating a -> AST Floating a
logexpAST1 (Free (Sig_log (Free (Sig_exp a)))) = a
\end{lstlisting}
This function has two syntactic warts.
First, the pattern is noisy.
To properly match the expression, we must alternate between the |Free| and |Sig_| constructors.
Second, the function is specialized to the |Floating| type class.
We would like our transformation to also works on subclasses of |Floating|.

We'll start by correcting the second problem using the |ViewPatterns| language extension.
|ViewPatterns| provides a syntactic sugar that lets us pattern match on the return value of a function.\footnote{
    The GHC wiki \cite{viewpatterns} provides details on exactly how the desugaring works.
}
We can use |ViewPatterns| to write polymorphic patterns by using polymorphic functions in the pattern.
Fortunately, we've already defined the exact function we need:
\begin{lstlisting}
unsafeExtractSig :: View alg1 alg2
    => Sig alg2 a -> Sig alg1 a
\end{lstlisting}
By using this function in the pattern,
we can write a polymorphic version of our program transformation.
\begin{lstlisting}
logexpAST2 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST2
    (Free (unsafeExtractSig -> Sig_log
    (Free (unsafeExtractSig -> Sig_exp a))))
    = a
\end{lstlisting}
This function now has the correct type signature (so we've solved problem 2),
but the pattern matches are even uglier (so problem 1 is worse).

The |PatternSynonyms| extension lets us define new patterns,
and we will use this feature to clean up our syntax.\footnote{
    Again, see the GHC Wiki \cite{patternsynonyms} for details on this language extension.
}
Specifically, we define a pattern synonym for each function contained in an FAlgebra.
Our naming convention is to prefix |AST_| to the name of the function.
These pattern synonyms combine the |Free| constructor with the view pattern from the previous example.
The synonyms for the |log| and |exp| functions are shown below.
\begin{lstlisting}
pattern AST_log :: View Floating alg
    => AST alg a -> AST alg a
pattern AST_log e
    <- Free (unsafeExtractSig -> Sig_log e)

pattern AST_exp :: View Floating alg
    => AST alg a -> AST alg a
pattern AST_exp e
    <- Free (unsafeExtractSig -> Sig_exp e)
\end{lstlisting}
Armed with these pattern synonyms,
we can rewrite our function in a much simpler form.
\begin{lstlisting}
logexpAST3 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST3 (AST_log (AST_exp a)) = a
\end{lstlisting}
This simple function works on our the motivating test case:
\begin{lstlisting}
ghci> logexpAST3 $ testFunc1 var1 :: AST Floating Var
var1
\end{lstlisting}
%stopzone
but it fails to work in this slightly more complex function:
\begin{lstlisting}
testFunc2 :: Floating a => a -> a
testFunc2 a = 1+exp(log a)
\end{lstlisting}
We haven't told our program transformations what to do when they don't pattern match!
The correct answer is that we want the transformations to recurse into the subexpressions,
applying the transformations there as well.
In the free monad, recursion is handled by a call to |fmap| and the base case is handled by extracting the value of |Pure|.
This gives our final version of the program transformation.
\begin{lstlisting}
logexpAST4 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST4 (AST_log (AST_exp a)) = a
logexpAST4 (Free f) = Free (fmap logexpAST4 f)
logexpAST4 (Pure a) = Pure a
\end{lstlisting}
Now we correctly recurse into subexpressions.
\begin{lstlisting}
ghci> logexpAST4 $ testFunc2 var1
    :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone
There's still one last wrinkle.
What if we need to apply the program transformation multiple times?
Consider the following test function.
\begin{lstlisting}
testFunc3 :: Floating a => a -> a
testFunc3 a = 1+log(log(log(exp(exp(exp a)))))
\end{lstlisting}
Applying our program transformation directly doesn't fix the function because the transformation needs to be applied repeatedly.
\begin{lstlisting}
ghci> logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+(log (log (exp (exp var1)))))
\end{lstlisting}
%stopzone
What we want is the \emph{fixed point} of the program transformation.
That is, we want to apply the program transformation repeatedly until it stops changing the AST.
The |fixAST| function below takes as input a transformation and returns the fixed point of that transformation.
\begin{lstlisting}
fixAST ::
    ( Eq (Sig alg (Free (Sig alg) a))
    , Eq a
    ) => (AST alg a -> AST alg a)
      -> (AST alg a -> AST alg a)
fixAST f ast = if ast==ast'
    then ast
    else fixAST f ast'
    where
        ast' = f ast
\end{lstlisting}
Now we can correctly modify the |testFunc3| function.
\begin{lstlisting}
ghci> fixAST logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone

@@

The transformation in in the previous section is relatively simple.
It could just as easily have been written using GHC's rewrite rules syntax \cite{jones2001playing}.
Rewrite rules are the primary mechanism for optimising source code in GHC.
One limitation of rewrite rules is that they cannot perform computations on the rewritten expression.
This limitation prevents rewrite rules from implementing the constant folding optimization.
Our program transformations, however, are powerful enough for this transformation.

@@

Here's an example of our constant folding in action.
First define a simple function.
\begin{lstlisting}
constFunc :: Num a => a -> a -> a
constFunc x1 x2 = x1*2+(7+2)*x2
\end{lstlisting}
If we call this function with two known parameters,
our |foldConstants| function is able to simplify the expression into a single constant.
\begin{lstlisting}
ghci> fixAST foldConstants (constFunc 2 3)
    :: AST Floating Double
(fromInteger 31)
\end{lstlisting}
Alternatively, we can call the function with two variable parameters and only fold the constants in the expression.
\begin{lstlisting}
ghci> fixAST foldConstants (constFunc var1 var2)
    :: AST Floating Var
((var1*(fromInteger 2))+((fromInteger 9)*var2))
\end{lstlisting}
We can even invoke the function with one known and one variable parameter.
\begin{lstlisting}
ghci> fixAST foldConstants (constFunc var1 3)
    :: AST Floating Var
((var1*(fromInteger 2))+(fromInteger 27))
\end{lstlisting}

@@

Our original motivation was to stabilize the |logLogistic1| function.
Stabilizing this function is a bit harder because it actually requires changing the type signature of the function by adding an |Ord| instance.
One of the requirements of a homogeneous function is that it have only a single class in the context.
To accommodate this condition, we'll create a new type class that has our desired types in the constraint.
\begin{lstlisting}
class (Floating a, Ord a) => FloatingOrd a
instance {-#OVERLAPPABLE#-} (Floating a, Ord a) => FloatingOrd a
\end{lstlisting}
Note that the |OVERLAPPABLE| pragma is required so that we can create an |Ord| instance for |AST FloatingOrd a|.

>>>>

\usepackage{amsmath,amsfonts}
\usepackage{hyperref}

@@

\begin{filecontents}{paper.bib}
@article{birkhoff1970heterogeneous,
  title={Heterogeneous algebras},
  author={Birkhoff, Garrett and Lipson, John D},
  journal={Journal of Combinatorial Theory},
  volume={8},
  number={1},
  pages={115--133},
  year={1970},
  publisher={Elsevier}
}

@article{goldberg1991,
  title={What every computer scientist should know about floating-point arithmetic},
  author={Goldberg, David},
  journal={ACM Computing Surveys (CSUR)},
  volume={23},
  number={1},
  pages={5--48},
  year={1991},
  publisher={ACM}
}

@@

@inproceedings{panchekha2015automatically,
  title={Automatically improving accuracy for floating point expressions},
  author={Panchekha, Pavel and Sanchez-Stern, Alex and Wilcox, James R and Tatlock, Zachary},
  booktitle={Proceedings of the 36th ACM SIGPLAN Conference on Programming Language Design and Implementation},
  pages={1--11},
  year={2015},
  organization={ACM}
}
@misc{viewpatterns,
  author = {GHC Wiki},
  title = {View patterns: lightweight views for Haskell},
  url = {https://ghc.haskell.org/trac/ghc/wiki/ViewPatterns},
  note = {Accessed: 2016-06-09}
}

@@

%\title{What's the point of homoiconicity?}
%\title{The (floating) point of homoiconicity}
\title{Homoiconicity Without Template Haskell}
%\subtitle{Functional Pearl}

@@

\begin{abstract}
We say a function is \emph{homoiconic} if it can be converted into an isomorphic abstract syntax tree (AST).
This AST can then be manipulated arbitrarily and converted back into a function.
Template Haskell gives a \emph{compile time} form of homoiconicity to Haskell programs;
but in this paper, we demonstrate another form of homoiconicity that relies on polymorphism and GHC's type programming capabilities.
To motivate our technique,
we construct a small library that automatically transforms numerically unstable code into stable code.

%This paper shows that many Haskell functions are \emph{homoiconic}.
%That is, we can convert these functions into isomorphic abstract syntax trees (ASTs),
%and these ASTs can be converted back into their original Haskell functions.
%%These ASTs are ordinary Haskell data types,
%%so they can be manipulated arbitrarily to perform program transformations.
%The construction does not require template Haskell,
%and so works on expressions available at either compile or run time.
%To motivate this homoiconicity,
%we construct a small library that automatically transforms numerically unstable code into stable code.

%Floating point numbers are dangerous.
%Due to their limited precision,
%seemingly correct mathematical formulae can give wildly inaccurate results.
%This paper shows that Haskell's type system can mitigate this danger.
%
%The main observation is that many numerical Haskell functions are \emph{homoiconic}.
%That is, we can easily convert them into isomorphic abstract syntax trees (ASTs).
%The construction does not use template Haskell, and so works at both compile and run time.
%We will use these ASTs to solve two problems with floating point numbers:
%\begin{enumerate}
%\item
%We will write higher order functions that automatically stabilize (or optimize!) mathematical formulae.
%%This lets one programmer write naive floating point code;
%%then an expert on numerical analysis can write code that automatically improves the naive code.
%\item
%We will specify type class laws for floating point arithmetic that can be automatically tested.
%Existing test frameworks do not work on floating point numbers because floats do not obey many traditional algebraic laws like associativity.
%\end{enumerate}
%We demonstrate these techniques using GHC's built-in numeric class hierarchy,
%but these techniques generalize to user defined class hierarchies as well.
\end{abstract}

@@

In this paper, we see how to write functions like |stabilize| that perform program transformations.
These functions follow the following pattern.
First, convert the input function into an abstract syntax tree (AST).
Next, manipulate the AST to remove the numerical instabilities.
Finally, convert the AST back into a callable function.
%The key observation is that certain Haskell functions are \emph{homoiconic}.
%That is, we can convert them into isomorphic abstract syntax trees (ASTs).
%The key observation is that certain Haskell functions (which we call \emph{homogeneous} and \emph{heterogeneous}) can be converted into equivalent abstract syntax trees (ASTs).
%The |stabilize| function converts its input function into an AST,
%then manipulate that AST to remove numerical stabilities,
%and finally convert the AST back into a callable function.

@@

There are two intuitions for homogeneous type classes.
First, they have ASTs that contain only a single type variable.
Second, we can construct an FAlgebra for these type classes.

@@

In this section we first define homogeneous type classes and functions.
Then we show that there is an FAlgebra for each homogeneous type class,
and that this FAlgebra makes homogeneous functions homoiconic.

@@

\begin{defn}
We call a function homogeneous if:
\begin{enumerate}
\item
there is exactly one type variable in the signature, which we denote by |a|;
\item
each of the function's parameters is either
\begin{enumerate}
\item
|a|, or
\item
concrete;
\end{enumerate}
\item
the function's return type is |a|; and
\item
the constraints contain exactly one homogeneous type classes applied to |a| and nothing else.
\end{enumerate}
The |logLogistic1| function is homogeneous,
and Figure \ref{code:ghc} shows many more examples taken from the Prelude's numeric hierarchy.
\end{defn}

%In the remainder of this section we describe a canonical method for constructing FAlgebras from homogeneous type classes and show that this construction makes homogeneous functions homoiconic.
%The construction is tedious to implement by hand,
%so the \homoiconic library provides the function |mkFAlgebra| to automate the boilerplate.
%This function is provided as a convenience and is not strictly necessary for the technique.

@@

The data family |Sig| encodes what mathematicians call the \emph{signature} of the FAlgebra.
The signature defines all the operations that can be performed on an FAlgebra.
Each |Sig| data instance can have many constructors,
and these constructors come in two flavors.
First is the \emph{function constructor}.
There should be one function constructor for each homogeneous class function, and
this constructor should have the same parameters as the function.
For the |Fractional| example, we've defined the three constructors |Sig_div|, |Sig_recip|, and |Sig_fromRational| for the three class methods |(/)|, |recip|, and |fromRational| respectively.
The second constructor flavor is the \emph{superclass constructor}.
There should be one superclass constructor for each superclass, and
this constructor should store the |Sig| of the corresponding superclass.
For the |Fractional| example, we have the |Sig_Fractional_Num| constructor corresponding to the |Num| superclass.

@@

\subsection{Constructing the AST}
\label{sec:hom.cons}

@@

Using our uniform representation of an AST,
we can make the |AST| type an instance of any homogeneous type class.
As an example, the |Fractional| instance is shown below.
\begin{lstlisting}
instance View Fractional alg
    => Fractional (AST alg a) where
    (/) e1 e2 = Free $ embedSig $ Sig_div e1 e2
    recip e   = Free $ embedSig $ Sig_recip e
    fromRational r
        = Free $ embedSig $ Sig_fromRational r
\end{lstlisting}
For each function in the type class,
we simply embed the function's constructor into the AST by prepending the expression |Free $ embedSig $|.

These instances let us easily create ASTs.
The |Num| AST can be created with the expression
\begin{lstlisting}
expr4 :: AST Num Double
expr4 = (1+2)+3
\end{lstlisting}
And the |Floating| AST can be created by just changing the type signature.
\begin{lstlisting}
expr5 :: AST Floating Double
expr5 = (1+2)+3
\end{lstlisting}
Notice that in these last two examples we did not embed the numbers into the AST with the |Pure| constructor.
Since |AST| is an instance of |Num|,
GHC will call |fromInteger| to embed the number for us automatically.

@@

\subsection{Showing functions}
Reconsider our |logLogistic1| function.
It has type
\begin{lstlisting}
logLogistic1 :: Floating a => a -> a
\end{lstlisting}
What happens if we specialize the |a| parameter to |AST Floating Double|?
We get a peek at the function's internals.
\begin{lstlisting}
ghci> logLogistic1 1 :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (fromInteger 1))))))
\end{lstlisting}
We don't just have to input single numbers;
we can also input full expressions.
These expressions will also get displayed,
intermixed with the original function.
\begin{lstlisting}
ghci> logLogistic1 (1*3+4) :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (((fromInteger 1)*(fromInteger 3))
    +(fromInteger 4)))))))
\end{lstlisting}
And we can even display the resulting AST of nested function calls.
\begin{lstlisting}
ghci> logLogistic1 (logLogistic1 1)
    :: AST Floating Double
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate (log ((fromInteger 1)/((fromInteger 1)
    +(exp (negate (fromInteger 1)))))))))))
\end{lstlisting}
It would be nice if the displayed function could show variables,
rather than just numbers.
To accomplish this, we need a variable type.
\begin{lstlisting}
newtype Var = Var String
\end{lstlisting}
Values of type |Var| cannot be plugged directly into the |logLogistic1| function because |Var| is not an instance of |Floating|.
Fortunately, we have a way to make any type an instance of |Floating| via |Floating|'s |AST| type.
More generically, we define the following variables.
\begin{lstlisting}
var1 :: AST alg Var
var1 = Pure "var1"

@@

var3 :: AST alg Var
var3 = Pure "var3"
\end{lstlisting}
And we can use them to recover the exact AST of a function.
\begin{lstlisting}
ghci> logLogistic1 var1 :: AST Floating Var
(log ((fromInteger 1)/((fromInteger 1)+(exp
    (negate var1)))))
\end{lstlisting}
At this point, you might be tempted to write a |Show| instance for functions.
\begin{lstlisting}
instance
    ( FAlgebra alg
    , Show (Sig alg (Free (Sig alg) Var))
    ) => Show (AST alg Var -> AST alg Var) where
    show f = show (f var1)
\end{lstlisting}
Unfortunately, if you try to use the instance naively
\begin{lstlisting}
ghci> logLogistic1
\end{lstlisting}
then ghci will complain about ambiguous type variables.
GHC's defaulting mechanism is not powerful enough for it to properly specialize the function without a type signature.
So while this construction does give us a convenient way to inspect the contents of a function,
it's not quite powerful enough for a proper |Show| instance.

@@

\subsection{Cheating with template Haskell}

For each homogeneous type class,
there are many boilerplate instances that we need to write.
The \homoiconic library provides a template Haskell function |mkFAlgebra| in the |Homogeneous.FAlgebra| module that generates the |FAlgebra|, |Functor|, |Show|, and |View| instances plus all of the pattern synonyms.

@@

We're now ready to write program transformations that manipulate the ASTs created in the previous section.
In this section, we'll first demonstrate how to pattern match on our ASTs using the |ViewPatterns| and |PatternSynonyms| language extensions.
%Then we'll see three simple examples of program transformations implemented in this way.
We'll also see three motivating examples of useful transformations on floating point operations.

@@

Consider the following simple,
but numerically unstable function.
\begin{lstlisting}
testFunc1 :: Floating a => a -> a
testFunc1 = log (exp a)
\end{lstlisting}
As in our |logLogistic1| function,
|testFunc| gives incorrect answers when |a| is specialized to |Double|.
Inputs greater than (approximately) 710 cause the |exp| function to overflow to infinity;
the log of infinity is still infinity;
so the returned answer is infinity.
But this wrong.
On the real numbers, the $\log$ function is defined to be the inverse of $\exp$,
so the final result should be exactly the same as the input.
We'll write a program transformation that stabilizes the |testFunc1| function,
then we'll demonstrate how the |ViewPatterns| and |PatternSynonyms| language extensions simplify this transformation.
\label{sec:viewpatterns}

Our goal is to take an input AST,
find all occurrences of the expression |log (exp x)|,
and replace them with just |x|.
(Here |x| can be an arbitrary AST, not just a single value.)
We will create a series of four functions to accomplish this task,
each improving on the previous.
Here is the first.
\begin{lstlisting}
logexpAST1 :: AST Floating a -> AST Floating a
logexpAST1 (Free (Sig_log (Free (Sig_exp a)))) = a
\end{lstlisting}
This function has two syntactic warts.
First, the pattern is verbose.
To properly match the expression, we must alternate between the |Free| and |Sig_| constructors.
It would be better to have a single constructor per operation.
Second, the function is specialized to the |Floating| type class.
We would like our transformation to also works on subclasses of |Floating|.

We'll start by correcting the second problem using the |ViewPatterns| language extension.
|ViewPatterns| provides a syntactic sugar that lets us pattern match on the return value of a function.\footnote{
    The GHC wiki \cite{viewpatterns} provides details on exactly how the desugaring works.
}
We can use |ViewPatterns| to write polymorphic patterns by using polymorphic functions in the pattern.
We need a polymorphic function for extracting a |Sig alg a| type from a |Sig Floating a| whenever |alg| is a subclass of |Floating|.
Fortunately, we've already defined such a function:
\begin{lstlisting}
unsafeExtractSig :: View alg1 alg2
    => Sig alg2 a -> Sig alg1 a
\end{lstlisting}
Using it in the pattern gives us
\begin{lstlisting}
logexpAST2 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST2
    (Free (unsafeExtractSig -> Sig_log
    (Free (unsafeExtractSig -> Sig_exp a))))
    = a
\end{lstlisting}
This modified program transformation now has the correct polymorphic type signature
(so we've solved problem 2),
but the pattern matches are even uglier
(so we've made problem 1 worse).

The |PatternSynonyms| extension lets us define new patterns,\footnote{
    Again, see the GHC Wiki \cite{patternsynonyms} for details on this language extension.
}
and we will use this feature to clean up our syntax.
Specifically, we define a pattern synonym for each function contained in an FAlgebra.
Our naming convention is to prefix |AST_| to the name of the function.
These pattern synonyms combine the |Free| constructor with the view pattern from the previous example.
The synonyms for the |log| and |exp| functions are shown below.
\begin{lstlisting}
pattern AST_log :: View Floating alg
    => AST alg a -> AST alg a
pattern AST_log e
    <- Free (unsafeExtractSig -> Sig_log e)

pattern AST_exp :: View Floating alg
    => AST alg a -> AST alg a
pattern AST_exp e
    <- Free (unsafeExtractSig -> Sig_exp e)
\end{lstlisting}
Armed with these pattern synonyms,
we can rewrite our function in a much simpler form.
\begin{lstlisting}
logexpAST3 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST3 (AST_log (AST_exp a)) = a
\end{lstlisting}
This simple function works on our the motivating test case:
\begin{lstlisting}
ghci> logexpAST3 $ testFunc1 var1 :: AST Floating Var
var1
\end{lstlisting}
%stopzone
but it fails on this slightly more complex function:
\begin{lstlisting}
testFunc2 :: Floating a => a -> a
testFunc2 a = 1+exp(log a)
\end{lstlisting}
The problem is that we haven't told our program transformations what to do when they don't pattern match!
When pattern matching fails, we want our transformations to recurse into the subexpressions,
applying the transformations there as well.
In the free monad, recursion is handled by a call to |fmap| and the base case is handled by extracting the contents of |Pure|.
This gives our final version of the program transformation.
\begin{lstlisting}
logexpAST4 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST4 (AST_log (AST_exp a)) = a
logexpAST4 (Free f) = Free (fmap logexpAST4 f)
logexpAST4 (Pure a) = Pure a
\end{lstlisting}
Now we correctly recurse into subexpressions.
\begin{lstlisting}
ghci> logexpAST4 $ testFunc2 var1
    :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone
But there's still one last wrinkle.
What if we need to apply the program transformation multiple times?
Consider the following test function.
\begin{lstlisting}
testFunc3 :: Floating a => a -> a
testFunc3 a = 1+log(log(log(exp(exp(exp a)))))
\end{lstlisting}
Applying our program transformation directly doesn't correct the function because the transformation needs to be applied repeatedly.
\begin{lstlisting}
ghci> logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+(log (log (exp (exp var1)))))
\end{lstlisting}
%stopzone
What we want is the \emph{fixed point} of the program transformation.
That is, we want to apply the program transformation repeatedly until it stops changing the AST.
The |fixAST| function below takes as input a transformation and returns the fixed point of that transformation.
\begin{lstlisting}
fixAST ::
    ( Eq (Sig alg (Free (Sig alg) a))
    , Eq a
    ) => (AST alg a -> AST alg a)
      -> (AST alg a -> AST alg a)
fixAST f ast = if ast==ast'
    then ast
    else fixAST f ast'
    where
        ast' = f ast
\end{lstlisting}
Now we can correctly modify the |testFunc3| function.
\begin{lstlisting}
ghci> fixAST logexpAST4 $ testFunc3 var1
    :: AST Floating Var
((fromInteger 1)+var1)
\end{lstlisting}
%stopzone
And that's all there is to writing a program transformation.
More complex transformations just involve more complex pattern matching.

@@

The transformation in in the previous section is relatively simple.
For example, it could have been implemented using GHC's rewrite rules \cite{jones2001playing}.
Rewrite rules are the primary mechanism for optimising source code in GHC,
but they are limited to transformations that involve no computation.
This limitation prevents rewrite rules from implementing a number of optimizations,
one of them being being constant folding.
Constant folding is the process of reducing an expression containing several constants into a single constant.
For example, converting the expression |(1+2)+3| into just the number |6|.
%(Due to this limitation, GHC has special ad-hoc code just for constant folding.)
Our program transformations, however, can implement arbitrary computation and so are powerful enough to handle constant folding.

@@

Let's see an example of constant folding in action.
First define a simple function with two parameters a handful of constants.
\begin{lstlisting}
func :: Num a => a -> a -> a
func x1 x2 = x1*2+(7+2)*x2
\end{lstlisting}
If we call |func| with two constant parameters,
our |foldConstants| function is able to simplify the resulting expression into a single constant.
\begin{lstlisting}
ghci> fixAST foldConstants (func 2 3)
    :: AST Floating Double
(fromInteger 31)
\end{lstlisting}
Alternatively, we can call the function with two variable parameters.
Only the constants in the original function will be folded.
\begin{lstlisting}
ghci> fixAST foldConstants (func var1 var2)
    :: AST Floating Var
((var1*(fromInteger 2))+((fromInteger 9)*var2))
\end{lstlisting}
And finally, we can call the function with one constant and one variable parameter.
\begin{lstlisting}
ghci> fixAST foldConstants (func var1 3)
    :: AST Floating Var
((var1*(fromInteger 2))+(fromInteger 27))
\end{lstlisting}

@@

Now let's see how to stabilize the |logLogistic1| function.
Stabilizing this function is a bit harder because it requires changing the type signature of the function---the constraints on |logLogistic1| require only a |Floating| instance, but the constraints on |logLogistic2| require both a |Floating| and |Ord| instance.
Recall that one of the requirements for a function to be homogeneous is that it have only a single class in the context.
Therefore the stabilized function |logLogistic2| is not homogeneous.
